Clustering based on probability distributions with application on residential customers

### 2021-10-20 main.Rmd
adding references

<!-- In contrast to models, a feature-based strategy is used to explicitly define or automatically extract the curves’ key time features, for instance by application of PCA on the daily curves [ -->

@@

The proposed methodology aim to leverage the intrinsic temporal data structure hidden in time series data. The foundation of our method is unsupervised clustering algorithms based exclusively on time-series features. First, we study the underlying distributions that may have resulted in different patterns across temporal granularities in order to identify a mechanism to classify them based on the similarity of those distributions. It is worth noting that when studying these similarities, a variety of objectives may be pursued. One objective could be to group consumers with similar shapes over all relevant cyclic granularities. In this scenario, the variation in customers within each group is in magnitude rather than shape, while the variation between groups is only in shape. Most clustering algorithms offer only daily energy profiles throughout the hours of the day, but we suggest a broader approach to the problem, aiming to group consumers with similar shapes across all significant cyclic granularities. Another purpose of clustering could be to group customers that have similar differences in patterns across all major cyclic granularities, capturing similar jumps across categories regardless of the overall shape. For example, in the first goal, similar shapes across hours of the day will be grouped together, resulting in customers with similar behaviour across all hours of the day, whereas in the second goal, any similar big-enough jumps across hours of the day will be clubbed together, regardless of which hour of the day it is. Both of these objectives may be useful in a practical context and, depending on the data set, may or may not propose the same customer classification.

Depending on the goal of clustering, the distance metric for defining similarity would be different. These distance metrics could be fed into a clustering algorithm to break large data sets into subgroups that can then be analyzed separately. These clusters may be commonly associated with real-world data segmentation. However, since the data is unlabeled a priori, more information is required to corroborate this. This section presents the work flow of the methodology:

@@

\noindent These cyclic granularities are useful for exploring repetitive patterns in time series data that get lost in the linear representation of time. It is advantageous to consider only those cyclic granularities across which there is a significant repetitive pattern for the majority of customers or noteworthy in an electricity-behavior context. In that case, when the customers are grouped, we can expect to observe some interesting patterns across the categories of the cyclic granularities considered. [XXX reference 2nd chapter] proposes a way to select significant cyclic granularities and harmonies which is used for this paper.

@@

(a) Parallel coordinate plot: Parallel coordinate plot (Inselberg 1985, Wegman 1990) are widely used to display high-dimensional and multivariate data, allowing visual grouping to detect patterns. In a Parallel Coordinates Plot, each variable has its own axis, which are all parallel. Each axis is connected by a series of lines. That is, each line is made up of connected points on each axis. The order of the axes can affect how the reader interprets the data. This is because adjacent variables are more easily perceived than non-adjacent variables and rearranging the axes can reveal patterns or correlations between variables. Scattered plots of the p variables are arranged in a scatterplot matrix. It's a neat way to show multiple relationships at once, and it allows us to compare all the plots at once.

@@

(d) Tour: A tour () is a collection of interpolated linear projections of multivariate data into a lower-dimensional space. The sequence is seen as a dynamic visualisation, enabling the viewer to observe the shadows cast by the high-dimensional data in a lower-dimensional view.
<!-- and also provides examples on how one might spin and brush using a tour, to build up a cluster analysis manually. -->

>>>>

<!-- In contrast to models, a feature-based strategy is used to explicitly define or automatically extract the curves’ key time features, for instance by application of PCA on the daily curves [ --

@@

<!-- Because of this importance, countless approaches to estimate time series similarity have been proposed.  -->

The proposed methodology aim to leverage the intrinsic temporal data structure hidden in time series data. The foundation of our method is unsupervised clustering algorithms based exclusively on the time-series data. The similarity measure is the most essential ingredient of time series clustering. The (dis) similarity measure in this paper focuses on looking at the (dis) similarity between underlying distributions that may have resulted in different patterns across different cyclic temporal granularities. It is worth noting that when studying these similarities, a variety of objectives may be pursued. One objective could be to group time series with similar shapes over all relevant cyclic granularities. In this scenario, the variation in customers within each group is in magnitude rather than shape, while the variation between groups is only in shape. There
are distance measures are used for shape-based clustering [Ding et al. 2008; Wang
et al. 2013] and many more but none of them look at the probability distributions while computing similarity. Moreover, most distance measures offer similar shape across just one dimension. For example, we often see "similar" daily energy profiles across hours of the day, but we suggest a broader approach to the problem, aiming to group consumers with similar distributional shape across all significant cyclic granularities. Another purpose of clustering could be to group customers that have similar differences in patterns across all major cyclic granularities, capturing similar jumps across categories regardless of the overall shape. For example, in the first goal, similar shapes across hours of the day will be grouped together, resulting in customers with similar behaviour across all hours of the day, whereas in the second goal, any similar big-enough jumps across hours of the day will be clubbed together, regardless of which hour of the day it is. Both of these objectives may be useful in a practical context and, depending on the data set, may or may not propose the same customer classification. Depending on the goal of clustering, the distance metric for defining similarity would be different. These distance metrics could be fed into a clustering algorithm to break large data sets into subgroups that can then be analyzed separately. These clusters may be commonly associated with real-world data segmentation. However, since the data is unlabeled a priori, more information is required to corroborate this. This section presents the work flow of the methodology:

@@

\noindent These cyclic granularities are useful for exploring repetitive patterns in time series data that get lost in the linear representation of time. It is advantageous to consider only those cyclic granularities across which there is a significant repetitive pattern for the majority of customers or noteworthy in an electricity-behavior context. In that case, when the customers are grouped, we can expect to observe some interesting patterns across the categories of the cyclic granularities considered. [@Gupta2021-hakear] proposes a way to select significant cyclic granularities and harmonies which is used for this paper.

@@

(a) Parallel coordinate plot: Parallel coordinate plot (@wegman1990hyperdimensional) are widely used to display high-dimensional and multivariate data, allowing visual grouping to detect patterns. In a Parallel Coordinates Plot, each variable has its own axis, which are all parallel. Each axis is connected by a series of lines. That is, each line is made up of connected points on each axis. The order of the axes can affect how the reader interprets the data. This is because adjacent variables are more easily perceived than non-adjacent variables and rearranging the axes can reveal patterns or correlations between variables. Scattered plots of the p variables are arranged in a scatterplot matrix. It's a neat way to show multiple relationships at once, and it allows us to compare all the plots at once.

@@

(d) Tour: A tour (@wickham2011tourr) is a collection of interpolated linear projections of multivariate data into a lower-dimensional space. The sequence is seen as a dynamic visualization, enabling the viewer to observe the shadows cast by the high-dimensional data in a lower-dimensional view.
<!-- and also provides examples on how one might spin and brush using a tour, to build up a cluster analysis manually. -->
